### 2016-04-14 Paper_v5.tex
Results and disucssion updates

\subsection*{DOPS minimized benchmark problems with a fixed number of function evaluations.}
First, we tested DOPS and four other meta-heuristic approaches on the minimization of the Ackley and Rastrigin functions.
DOPS performed similarly or outperformed the other meta-heuristic approaches on both of these test functions (Fig. \ref{fig-testfunctions}).
The Ackley and Rastrigin functions both have multiple local extrema and attain a global minimum value of zero.
In each case, we fixed the maximum number of function evaluations at 4000 and ran 25 independent experiments with different initial parameter vectors.
DOPS found optimal or near optimal solutions for both the 10-dimensional Ackley (Fig. \ref{fig-testfunctions}A) and Rastrigin (Fig. \ref{fig-testfunctions}B) functions within the budget of function evaluations.
In each of the 10-dimensional cases, other meta-heurtistics such as DDS and DE also performed well, however DOPS consistently outperformed all other approaches tested.
This performance difference was more pronounced as the dimension of the search problem increased; for a 300-dimensional Rastrigin function DOPS was the only approach to find an optimal or near optimal solution
within the function evaluation budget (Fig. \ref{fig-testfunctions}B).
Taken together, these results suggested that DOPS could perform at least as well as other meta-heuristic approaches on small test problems, but was especially suited to large dimension search spaces.
Next, we tested DOPS on benchmark biochemical models of varying complexity.

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked ordered these example problems from most expensive to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
The biochemical reactions were modeled using modular rate laws and the generalized form of Michaelis-Menten kinetics.
The expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1).
In both cases, synthetic time series data sets were generated using known parameter values.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at 4000 for DOPS and trained both the models against the synthetic experimental data.
In both cases we found good fits  (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}) to the problems within 4000 evaluations.
We recaptured the nominal parameter values within 4000 evaluations (Fig. \ref{fig-benchmark}).

@@

\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
We compared the performance of DOPS with simulated annealing (SA), differential evolution (DE), and dynamically dimensioned search (DDS) for a model of the human coagulation cascade.
Coagulation is an archetype biochemical network that is highly interconnected, containing both positive and negative feedback (Fig. \ref{fig-coagulation-network}).
The biochemistry of coagulation, though complex, has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model},
and reliable experimental protocols have been developed to interrogate the system \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
Thus, the human coagulation cascade is an ideal test case;
coagulation is challenging because it contains both fast and slow dynamics, but also accessible because of the availability of comprehensive data sets for model identification and validation.
In this study, we used the coagulation model of Luan et al \cite{luan2007computationally}. Luan and coworkers modeled coagulation as a system of coupled non-linear ordinary differential equations.
The Luan model contained 148 parameters and 92 species and has been validated using 21 published datasets.

Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets.
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination [13,14].
Initiation requires a trigger event, for example a vessel injury which exposes tissue factor (TF), which leads to the activation of factor VII (FVIIa) and the formation of the TF/FVIIa complex.
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal.
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation.
Tissue Factor Pathway Inhibitor (TFPI) inhibits upstream activation events, while antithrombin III (ATIII) neutralizes several of the proteases generated during coagulation, including thrombin.
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR), converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of amplification complexes.
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.

%Coagulation is regulated by a set of serine proteases known as coagulation factors and a type of blood cell called a platelet.
%In the circulation, coagulation factors are generally in an inactive state known as a zymogen.
%These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.  For example when coagulation is initiated through the tissue factor pathway, tissue factor and activated factor VIIa (FVIIa) form a complex that activates factors X (fX) and IX (fIX). Activated factor X (fXa) thereafter activates downstream factors VIII and IX. The initial activation leads to the production of picomolar amounts of thrombin (fIIa) which activates platelets and amplifies its own production through the formation of a prothrombinase complex (FXa-FVa) on the surface of the activated platelets. Thrombin also  downregulates its own production by forming a complex with thrombomodulin which then activates protein C (PC). PC inhibits the formation of prothrombinase complex. In addition, Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner. Antithrombin III (ATIII)  inhibits all proteases. surface protein thrombomodulin (TM).
%This choice of using a linear combination of two different error functions was motivated by poor validation results while using a single error function.

To estimate the coagulation model parameters, we used data sets from TF/VIIa initiated coagulation in the absence of anticoagulants.
The objective function was a linear combination of two error functions, representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
We restricted the number of function evaluations to $\mathcal{N}$ = 4000 for each algorithm we tested, and performed 25 trials of each experiment to collect average performance data (Table ZZ).DOPS converged faster and had a lower final error compared to the other algorithms (Fig. \ref{fig-convergence}).
Within the first 1000 function evaluations of DOPS, there was a rapid drop in error.
Approximately between 500-1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varied from trial to trial since the switch was based on the error from the swarm phase). Overall at the end of 4000 function evaluations DOPS minimized the error (final objective error is 5.5916e+04) to a greater extent than the other algorithms.

Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Fig.\ref{fig-train}).
The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
Subsequently we used these optimal parameters to make model predictions that was compared against completely 'unseen' or untrained experimental data where
coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Fig.\ref{fig-validation}).

\subsection*{Influence of the number of sub-swarms on DOPS performance.}
We attempted to quantify the influence of sub-swarms and DDS phase on the error convergence rates. To do so we considered only the multi swarm search without the DDS phase for 4000 function evaluations and 40 particles on the coagulation model. While doing the multi swarm search without DDS we varied the number of sub-swarms as one, two, four, five and eight. The number of sub-swarms were chosen so that each swarm contained the same number of particles. Hence we did not consider swarm numbers of three and seven. All the other parameters remained the same for each of these searches. We observed that the higher sub-swarm numbers had faster convergence rates with eight having the fastest rate. However this difference in convergences is not pronounced amongst four, five and eight. This suggests that there is an optimal number of particles per swarm beyond which there is no significant advantage.

We also observe that when these convergence rates are compared with DOPS, DOPS shows a very rapid drop when the swarm searches begin to saturate. The green region indicates when swarm search switches to a DDS based search. As we see in (Fig. ) the error convergence rates tend to remain mostly flat for the purely swarm searches with different swarm numbers. However in DOPS as soon as convergence rate begins to saturate it automatically switches to a DDS phase which leads to a more pronounced drop. Thus this automated switching strategy appears to be crucial in leading the algorithm out of local optima or saddle point regions.

\clearpage
\section*{Materials and Methods}

@@

\clearpage
\section*{Discussion}
Our study presents a novel approach for high-dimensional parameter estimation in complex biological systems with relatively few function evaluations. In this approach we combined a variant of a well known meta heuristic particle swarm optimization with Dynamically Dimensioned Search (DDS). We tested our approach on an ODE model of coagulation with 148 parameters and 92 species. Coagulation is an ideal system to test our approach since the biology is well known and complex, with multiple feed back loops that are tightly regulated. We used experimental data under different conditions to obtain optimal parameters and used these parameters to make predictions against unseen experimental data. We obtained good fits and made sufficiently accurate enough predictions using parameters obtained from 4000 function evaluations. Further, we also used high-dimensional forms of commonly used test functions of global optimization and showed that we were able to find the global minimum for 300 dimensional Ackley and Rastrigin functions faster than other meta-heuristics. We also considered two recently published benchmark problems to test parameter optimization approaches and showed that we were able to retrieve the nominal parameter vector within 4000 function evaluations.
Meta-heuristic approaches are generally effective in finding close to optimum solutions of complex, multi-modal functions. In addition, they generally obviate the need for any \textit{a priori} knowledge (like function derivative). However they take an exorbitant number of objective function evaluations to come close to an optimum. When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations. As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.  In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.  Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations. Though their approach is based on a single solution, the decision vector carries forward from one iteration to the next. Hence it behaves similar to a population based search. In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization. Particle Swarm Optimization (PSO) is a population based meta heuristic which does not have any complex operations like recombination, mutation or selection that are associated with other population based meta-heuristics like Differential Evolution (DE) or Genetic Algorithm (GA). Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters. Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}. Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.  Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.

We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search. Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution. Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone. The computational cost was prohibitively high for parameter estimation on this problem using a standard PSO. Thus we used SA, DE and DDS for the purpose of comparison.
 Choosing the number of function evaluations is largely a function of cost and complexity of the objective function. Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold. However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion. In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case. We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}. In this work by Villaverde et al. they used a population based search enhanced Scatter Search (eSS) to estimate the biochemical model parameters. Quite surprisingly we took a couple of orders lesser number of function evaluations 10\textsuperscript{3} to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 10\textsuperscript{5} number of evaluations.  The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.

A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem. In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase. We used the same parameters for all the problems. The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing. For SA, we optimized the cooling schedule for the coagulation model. Thus, in this approach any overhead that usually comes with additional function evaluations to pre optimize parameters was avoided.

The performance of our approach seems impressive given that it performed well on different complex systems with no pre optimization of algorithm parameters being required. We comfortably outperformed existing, widely used meta-heuristics and were also able to find minima of high dimensional global optimization test functions. Thus this approach may be well suited to large scale global optimization. In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared to enhanced Scatter Search (eSS). However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems. This aspect is currently beyond the scope of this study. Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions. In addition, the current implementation of the algorithm is designed to switch only once from the swarm phase to the DDS. In the DDS phase, the search uses only one candidate vector although there is a provision to start with a population of candidate vectors. Incorporating a more intelligent switching strategy that can do switch from swarm phase to DDS phase multiple times and having a population of candidate vectors are some aspects of the algorithm that can be studied further.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\textwidth]{./figs/Figure_6_TestFunctions-crop.pdf}
\caption{Performance of DOPS and other meta-heuristics for the Ackley and Rastrigin functions.
\textbf{A}: Mean scaled error versus the number of function evaluations for the 10-dimensional Ackley function. DOPS, DDS and DE find optimal or near optimal solutions within the specified number of function evaluations.
\textbf{B}: Mean scaled error versus the number of function evaluations for the 10-dimensional Rastrigin function. DOPS and DDS find optimal or near optimal solutions within the specified number of function evaluations.
\textbf{C}: Mean scaled error versus the number of function evaluations for the 300-dimensional Rastrigin function. DOPS is the only algorithm that finds an optimal or near optimal solution within the specified number of function evaluations. In all cases, the maximum number of function evaluations was 4000. Mean and standard deviation were calculated over 25 trials. }\label{fig-testfunctions}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Figure_2_Coagulation_v4.pdf}
\caption{Schematic of the extrinsic and intrinsic coagulation cascade\cite{luan2007computationally}. Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF)  following vessel injury. Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX). FXa activates downstream factors including factor VIII (fVIII) and fIX. Factor V (fV) is primarily activated by thrombin (FIIa). In addition, we included a secondary fV activation route involving FXa. FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa. FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.  Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation. Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner. Antithrombin III (ATIII)  inhibits all proteases. Thrombin inhibits itself binding the surface protein thrombomodulin (TM). The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa. }\label{fig-coagulation-network}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure_3_Errors_convergence_v2.pdf}
\caption{Error convergence rates of the five different algorithms on the coagulation model. The objective error is the mean over $N$= 25 trials. DOPS, DDS and SA have the steepest drop in error during first 300 function evaluations. Thereafter the error drop in DDS and SA remains nearly constant whereas DOPS continues to drops further. At the end of 4000 function evaluations DOPS attains the lowest error. The next best estimate using DDS is nearly 3 times greater than the lowest error using DDS.
}\label{fig-convergence}
\end{figure}

>>>>

\subsection*{DOPS minimized benchmark problems with a fixed number of function evaluations.}
We tested DOPS and four other meta-heuristic approaches on the minimization of the Ackley and Rastrigin functions.
DOPS performed similarly or outperformed the other meta-heuristic approaches on both of these test functions (Fig. \ref{fig-testfunctions}).
The Ackley and Rastrigin functions both have multiple local extrema and attain a global minimum value of zero.
In each case, we fixed the maximum number of function evaluations at $\mathcal{N}$ = 4000 and ran 25 independent experiments with different initial parameter vectors.
DOPS found optimal or near optimal solutions for both the 10-dimensional Ackley (Fig. \ref{fig-testfunctions}A) and Rastrigin (Fig. \ref{fig-testfunctions}B) functions within the budget of function evaluations.
In each of the 10-dimensional cases, other meta-heurtistics such as DDS and DE also performed well, however DOPS consistently outperformed all other approaches tested.
This performance difference was more pronounced as the dimension of the search problem increased; for a 300-dimensional Rastrigin function DOPS was the only approach to find an optimal or near optimal solution
within the function evaluation budget (Fig. \ref{fig-testfunctions}B).
Taken together, DOPS performed at least as well as other meta-heuristic approaches on small test problems, but was especially suited to large dimension search spaces.
Next, we tested DOPS on benchmark biochemical models of varying complexity.

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked ordered these example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters.
The biochemical reactions were modeled using modular rate laws and the generalized form of Michaelis-Menten kinetics.
The expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1). In both cases, synthetic time series data sets were generated using known parameter values, these data were then used by the parameter estimation methods.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
In both cases we found good fits  (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}) to the problems.
We also recaptured the nominal parameter values within the specified number of function evaluations (Fig. \ref{fig-benchmark}).
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data.
Next, we compared the performance of DOPS with the four other meta-heuristic approaches for a model of the human coagulation cascade.

@@

\subsection*{DOPS estimated the parameters of a human coagulation model.}
Coagulation is an archetype biochemical network that is highly interconnected, containing both negative and positive feedback (Fig. \ref{fig-coagulation-network}).
The biochemistry of coagulation, though complex, has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model},
and reliable experimental protocols have been developed to interrogate the system \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets.
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination [13,14].
Initiation requires a trigger event, for example a vessel injury which exposes tissue factor (TF), which leads to the activation of factor VII (FVIIa) and the formation of the TF/FVIIa complex. Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal.
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation.
Tissue Factor Pathway Inhibitor (TFPI) inhibits upstream activation events, while antithrombin III (ATIII) neutralizes several of the proteases generated during coagulation, including thrombin.Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR), converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of amplification complexes.
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.
Thus, the human coagulation cascade is an ideal test case;
coagulation is challenging because it contains both fast and slow dynamics, but also accessible because of the availability of comprehensive data sets for model identification and validation. In this study, we used the coagulation model of Luan et al \cite{luan2007computationally}, which is a coupled system of non-linear ordinary differential equations where biochemical interactions are modeled using mass action kinetics. The Luan model contained 148 parameters and 92 species and has been validated using 21 published experimental datasets.

%Coagulation is regulated by a set of serine proteases known as coagulation factors and a type of blood cell called a platelet.
%In the circulation, coagulation factors are generally in an inactive state known as a zymogen.
%These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.  For example when coagulation is initiated through the tissue factor pathway, tissue factor and activated factor VIIa (FVIIa) form a complex that activates factors X (fX) and IX (fIX). Activated factor X (fXa) thereafter activates downstream factors VIII and IX. The initial activation leads to the production of picomolar amounts of thrombin (fIIa) which activates platelets and amplifies its own production through the formation of a prothrombinase complex (FXa-FVa) on the surface of the activated platelets. Thrombin also  downregulates its own production by forming a complex with thrombomodulin which then activates protein C (PC). PC inhibits the formation of prothrombinase complex. In addition, Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner. Antithrombin III (ATIII)  inhibits all proteases. surface protein thrombomodulin (TM).
%This choice of using a linear combination of two different error functions was motivated by poor validation results while using a single error function.
%Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data
%The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
%Subsequently we used these optimal parameters to make model predictions that was compared against completely 'unseen' or untrained experimental data where
%coagulation was  respectively .

DOPS estimated the parameters of a human coagulation model for TF/VIIa initiated coagulation without anticoagulants (Fig.\ref{fig-train}).
The objective function was an unweighted linear combination of two error functions,
representing coagulation initiated with different concentrations of TF/FVIIa (5pM, 5nM) \cite{hockin2002model}.
We restricted the number of function evaluations to $\mathcal{N}$ = 4000 for each algorithm we tested, and performed 25 trials of each experiment to collect average performance data (Table \ref{table:benchmark-problems}). DOPS converged faster and had a lower final error compared to the other algorithms (Fig. \ref{fig-convergence}).
Within the first 25\% of function evaluations, DOPS produced a rapid drop in error followed by a slower but steady decline.
Approximately between 500-1000 function evaluations DOPS switched to the dynamically dimensioned search phase, however this switch iteration varied from trial to trial since the switch was based on the error from the swarm phase. Overall, DOPS minimized the error to a greater extent than the other meta-heuristic approaches.
However, it was unclear if the parameters obtained by DOPS had predictive power on unseen data. To address this question, we used the final parameters estimated by
DOPS to simulate data that was not used in training (coagulation initiated with 500pM,50pM,10pM TF/VIIa).
The optimal or near optimal parameters obtained by DOPS predicted unseen coagulation datasets (Fig.\ref{fig-validation}).
Taken together, DOPS estimated parameter sets with predictive power on unseen coagulation data using fewer function iterations than other meta-heuristics.
Next, we explored how the number of sub-swarms and the switch to DDS during the initial phase of an optimization trial influenced the performance of the approach.

\subsection*{The number of sub-swarms and phase switching influenced DOPS performance.}
A differentiating feature of DOPS compared to other meta-heuristics was the rapid error drop during the first 25\% of an optimization trial.
This drop occurred during the swarm phase and was punctuated by the switch to the dynamically dimensioned search phase of the approach.
We quantified the influence of the number of sub-swarms and switch to the DDS phase on error convergence by comparing DOPS with and without DDS for different numbers of sub-swarms (Fig. \ref{fig-sub-swarms}).
We considered only the multi swarm search without the DDS phase for $\mathcal{N}$ = 4000 function evaluations on the coagulation model.
While doing the multi swarm search without DDS, we used one, two, four, five and eight sub-swarms, with a total of 40 particles dived evenly amongst the swarms.
Hence we did not consider swarm numbers of three and seven.
All the other parameters remained the same for each of these searches.
Generally, the higher sub-swarm numbers had faster convergence.
However, the difference in convergences was not pronounced amongst four, five and eight, suggesting
there was an optimal number of particles per swarm beyond which there was no significant advantage.
We also observed that convergence rates are compared with DOPS, DOPS shows a very rapid drop when the swarm searches begin to saturate.
The green region indicates when swarm search switches to a DDS based search.
The error convergence rates tend to remain mostly flat for the purely swarm searches with different swarm numbers.
However in DOPS as soon as convergence rate begins to saturate it automatically switches to a DDS phase which leads to a more pronounced drop.
Thus, this automated switching strategy appears to be crucial in leading the algorithm out of local optima or saddle point regions.

\section*{Discussion}
In this study, we developed dynamic optimization with particle swarms (DOPS), a novel meta-heuristic for parameter estimation in models of biological systems.
DOPS combined multi-swarm particle swarm optimization, a global search approach, with the greedy strategy of dynamically dimensioned search to estimate optimal or nearly optimal solutions in a fixed number of function evaluations. We tested the performance of DOPS and four widely used meta-heuristics on the Ackley and Rastrigin test functions, a set of biochemical benchmark problems and a model of the human coagulation cascade.
As the number of parameters increased, DOPS outperformed the other meta-heuristics, generating optimal or nearly optimal solutions using significantly fewer function evaluations compared with the other methods. We tested the solutions generated by DOPS by comparing the estimated and true parameters in the benchmark studies, and by using the coagulation model to predict unseen experimental data. For both benchmark problems, DOPS retrieved the true parameters in significantly fewer function evaluations than other meta-heuristics.
For the coagulation model, we used experimental coagulation measurements under two different conditions to estimate optimal or nearly optimal parameters.
These parameters were then used to predict unseen coagulation data;
the coagulation model parameters estimated by DOPS predicted the correct thrombin dynamics following TF/FVIIa induced coagulation without anticoagulants.
Lastly, we showed the average performance of DOPS improved when combined with dynamically dimensioned search phase, compared to an identical multi-swarm approach alone.
Taken together, DOPS is a promising meta-heuristic for the estimation of parameters in large biochemical models.

Meta-heuristics are effective techniques to estimate optimal or nearly optimal solutions for complex, multi-modal functions.
However, they often take an exorbitant number of objective function evaluations to come close to an optimum. When the objective function evaluations tend to become expensive it is infeasible to take up a large number of evaluations. As the dimensionality of parameter space increases, the search region gets widened and thus the problem becomes more challenging.  In addition, most of these approaches require optimization of  'algorithm parameters' before the actual optimization and also involve computationally expensive update operations.  Tolson and Shoemaker, through DDS, showed that randomly perturbing a subset of dimensions in high dimensional parameter space is an effective way to obtain near optimal solutions with few function evaluations. Though their approach is based on a single solution, the decision vector carries forward from one iteration to the next. Hence it behaves similar to a population based search. In our approach, we tried to eliminate the probability of starting from a bad region of search using a variant of particle swarm optimization. Particle Swarm Optimization (PSO) is a population based meta heuristic which does not have any complex operations like recombination, mutation or selection that are associated with other population based meta-heuristics like Differential Evolution (DE) or Genetic Algorithm (GA). Several particle swarm variants have been proposed to improve the search ability and rate of convergence, that involve different neighborhood structures, multi-swarms or adaptive parameters. Multi-swarm PSO with small particle neighborhoods have been shown to better in searching on complex multi-modal solutions \cite{zhao2008dynamic}. Multi swarm methods, in addition, avoid rapid convergence to a local optimum or stable point and are able to generate diverse solutions.  Generation of diverse solutions in the early stage gives a better exploratory capability and thus converge of upon multiple optima.

We utilized this capacity of multi swarms to generate diverse candidate solutions which can be used as initial solutions for a DDS like search. Thus the solution vectors obtained at the end of multi swarm search phase have a better propensity to not get stuck in bad regions in DDS phase as opposed to starting with a totally random initial solution. Given a fixed number of function evaluations, we have been able to show that we were able to obtain better solutions for the coagulation model and the test functions faster that other commonly known meta-heuristics and also DDS alone. The computational cost was prohibitively high for parameter estimation on this problem using a standard PSO. Thus we used SA, DE and DDS for the purpose of comparison. Choosing the number of function evaluations is largely a function of cost and complexity of the objective function. Traditionally the stopping condition for a parameter optimization problem can be the number of function evaluations, percentage of initial error achieved or an absolute error threshold. However in case of complex, expensive functions where we desire a value within a certain period of time, the number of evaluations are used as a stopping criterion. In our current study we used a value of 4000 which we based upon the time taken (approximately 8-10 seconds) for a single objective function evaluation in the coagulation case. We used the same value of 4000 for benchmarks published by Villaverde and co-workers \cite{villaverde2015biopredyn}. In this work by Villaverde et al. they used a population based search enhanced Scatter Search (eSS) to estimate the biochemical model parameters. Quite surprisingly we took a couple of orders lesser number of function evaluations 10\textsuperscript{3} to obtain the optimal parameter vector as compared to the enhanced Scatter Search (eSS) with a local optimizer, which took around the order of 10\textsuperscript{5} number of evaluations.  The amount of CPU time taken (on an Intel Xeon processor 2.4 GHz) is lesser as compared to eSS on a similar architecture.
A surprisingly remarkable aspect about our algorithm has been that we did not 'pre optimize' any parameters of the algorithm to suit a specific problem. In the swarm phase this includes the number of particles, number of sub-swarms, acceleration constants or the number of generations after which the particles are redistributed and the neighborhood perturbation parameter in DDS phase. We used the same parameters for all the problems. The same rule was applied to the rest of the meta-heuristics barring Simulated Annealing. For SA, we optimized the cooling schedule for the coagulation model. Thus, in this approach any overhead that usually comes with additional function evaluations to pre optimize parameters was avoided.

The performance of DOPS was impressive given that it performed well on different complex systems with no pre optimization of algorithm parameters being required. We comfortably outperformed existing, widely used meta-heuristics and were also able to find minima of high dimensional global optimization test functions. Thus this approach may be well suited to large scale global optimization. In addition, surprisingly, we were able to obtain optimal parameter vectors for two different large scale systems biology models with a couple of orders fewer number of function evaluations as compared to enhanced Scatter Search (eSS). However it is quite possible that highly optimized versions of common meta heuristics may outperform us on these systems. This aspect is currently beyond the scope of this study. Our approach can also be combined with local derivative based searches to improve upon the accuracy of the solutions. In addition, the current implementation of the algorithm is designed to switch only once from the swarm phase to the DDS. In the DDS phase, the search uses only one candidate vector although there is a provision to start with a population of candidate vectors. Incorporating a more intelligent switching strategy that can do switch from swarm phase to DDS phase multiple times and having a population of candidate vectors are some aspects of the algorithm that can be studied further.

\section*{Materials and Methods}

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[width=0.7\textwidth]{./figs/Figure_2_TestFunctions-crop.pdf}
\caption{Performance of DOPS and other meta-heuristics for the Ackley and Rastrigin functions.
\textbf{A}: Mean scaled error versus the number of function evaluations for the 10-dimensional Ackley function. DOPS, DDS and DE find optimal or near optimal solutions within the specified number of function evaluations.
\textbf{B}: Mean scaled error versus the number of function evaluations for the 10-dimensional Rastrigin function. DOPS and DDS find optimal or near optimal solutions within the specified number of function evaluations.
\textbf{C}: Mean scaled error versus the number of function evaluations for the 300-dimensional Rastrigin function. DOPS is the only algorithm that finds an optimal or near optimal solution within the specified number of function evaluations. In all cases, the maximum number of function evaluations was 4000. Mean and standard deviation were calculated over 25 trials. }\label{fig-testfunctions}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Figure_3_Coagulation_v4.pdf}
\caption{Schematic of the extrinsic and intrinsic coagulation cascade\cite{luan2007computationally}. Inactive zymogens upstream (grey) are activated by exposure to tissue factor (TF)  following vessel injury. Tissue factor and activated factor VIIa (FVIIa) form a complex that activates factor X (fX) and IX (fIX). FXa activates downstream factors including factor VIII (fVIII) and fIX. Factor V (fV) is primarily activated by thrombin (FIIa). In addition, we included a secondary fV activation route involving FXa. FXa and FVa form a complex (prothrombinase) on activated platelets that converts prothrombin (fII) to FIIa. FIXa and FVIIIa can also form a complex (tenase) on activated platelets which catalyzes FXa formation.  Thrombin also activates upstream coagulation factors, forming a strong positive feedback ensuring rapid activation. Tissue factor pathway inhibitor (TFPI) downregulates FXa formation and activity by sequestering free FXa and TF-FVIIa in a FXa-dependent manner. Antithrombin III (ATIII)  inhibits all proteases. Thrombin inhibits itself binding the surface protein thrombomodulin (TM). The IIa-TM complex catalyzes the conversion of protein C (PC) to activated protein C (APC), which attenuates the coagulation response by the proteolytic cleavage of fV/FVa and fVIII/FVIIIa. }\label{fig-coagulation-network}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure_4_Errors_convergence_v2.pdf}
\caption{Error convergence rates of the five different algorithms on the coagulation model. The objective error is the mean over $N$= 25 trials. DOPS, DDS and SA have the steepest drop in error during first 300 function evaluations. Thereafter the error drop in DDS and SA remains nearly constant whereas DOPS continues to drops further. At the end of 4000 function evaluations DOPS attains the lowest error. The next best estimate using DDS is nearly 3 times greater than the lowest error using DDS.
}\label{fig-convergence}
\end{figure}
